```{r 08_07_setup, include=FALSE}
# Global knitr options - controls all code chunks
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)

# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(vegan)     # For CCA analysis
library(gridExtra) # For arranging multiple plots
library(RColorBrewer) # For color palettes
library(patchwork)  # For combining plots
library(ggrepel)   # For non-overlapping text labels

# Load global configuration
source("../R/global_config.R")

# Load data
load("../data/all_observations_split.RData")
load("../data/all_land_cover_data.RData")

# --- Canonical Correspondence Analysis (CCA) for 100m Land Cover Data ---

# Function to prepare species data matrix for CCA analysis by region
prepare_species_matrix <- function(region_data) {
  non_species_cols <- c("plot", "id_beach", "region")
  species_cols <- setdiff(names(region_data), non_species_cols)
  
  species_by_beach <- region_data %>%
    mutate(across(all_of(species_cols), bb_to_percentage)) %>%
    group_by(id_beach) %>%
    summarise(across(all_of(species_cols), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
    arrange(id_beach)
  
  species_matrix <- species_by_beach %>%
    select(all_of(species_cols)) %>%
    as.matrix()
  
  species_present <- colSums(species_matrix) > 0
  species_matrix_final <- species_matrix[, species_present, drop = FALSE]
  
  return(list(
    matrix = species_matrix_final,
    beach_ids = species_by_beach$id_beach
  ))
}

# Function to prepare environmental data for 100m analysis
prepare_env_matrix_100m <- function(land_cover_region_data, region_species_data) {
  beach_ids <- region_species_data$id_beach
  
  env_data <- land_cover_region_data %>%
    filter(id_beach %in% beach_ids) %>%
    arrange(id_beach)
  
  # Select 100m land cover variables
  land_cover_cols <- grep("^x100m_.*_percent$", names(env_data), value = TRUE)
  
  env_matrix <- env_data %>%
    select(id_beach, all_of(land_cover_cols)) %>%
    select(where(~ var(., na.rm = TRUE) > 0.01))
  
  # Clean column names
  numeric_cols <- setdiff(names(env_matrix), "id_beach")
  names(env_matrix)[names(env_matrix) %in% numeric_cols] <- 
    gsub("^x100m_(.*)_percent$", "\\1", numeric_cols)
  names(env_matrix) <- gsub("_", " ", names(env_matrix))
  names(env_matrix) <- tools::toTitleCase(names(env_matrix))
  names(env_matrix)[names(env_matrix) == "Id Beach"] <- "id_beach"
  
  return(env_matrix)
}

# Function to run CCA analysis at 100m scale
run_cca_analysis_100m <- function(region_data, land_cover_region_data, region_name) {
  cat("Running CCA analysis (100m scale) for", region_name, "...\n")
  
  if(nrow(region_data) < 3) {
    warning(paste("Not enough data for CCA analysis in", region_name))
    return(NULL)
  }
  
  species_result <- prepare_species_matrix(region_data)
  species_matrix <- species_result$matrix
  species_beach_ids <- species_result$beach_ids
  
  if(ncol(species_matrix) < 2) {
    warning(paste("Not enough species for CCA analysis in", region_name))
    return(NULL)
  }
  
  env_matrix <- prepare_env_matrix_100m(land_cover_region_data, region_data)
  env_matrix_matched <- env_matrix %>% 
    filter(id_beach %in% species_beach_ids) %>%
    arrange(id_beach)
  
  common_beaches <- intersect(species_beach_ids, env_matrix_matched$id_beach)
  
  if(length(common_beaches) < 3) {
    warning(paste("Not enough matching beaches for CCA analysis in", region_name))
    return(NULL)
  }
  
  beach_order <- sort(common_beaches)
  species_order <- match(beach_order, species_beach_ids)
  species_matrix_final <- species_matrix[species_order, , drop = FALSE]
  
  env_matrix_final <- env_matrix_matched %>% 
    filter(id_beach %in% beach_order) %>%
    arrange(id_beach)
  
  env_vars <- env_matrix_final %>% select(-id_beach)
  env_vars <- env_vars[, sapply(env_vars, var, na.rm = TRUE) > 0, drop = FALSE]
  
  if(ncol(env_vars) == 0) {
    warning(paste("No environmental variables with variation found in", region_name))
    return(NULL)
  }
  
  tryCatch({
    cca_result <- cca(species_matrix_final ~ ., data = env_vars)
    eigenvals <- eigenvals(cca_result)
    total_inertia <- sum(eigenvals)
    constrained_inertia <- sum(eigenvals[1:min(length(eigenvals), ncol(env_vars))])
    explained_variance <- constrained_inertia / total_inertia * 100
    
    return(list(
      cca = cca_result,
      species_matrix = species_matrix_final,
      env_matrix = env_vars,
      explained_variance = explained_variance,
      beach_ids = beach_order,
      region = region_name
    ))
  }, error = function(e) {
    warning(paste("CCA analysis failed for", region_name, ":", e$message))
    return(NULL)
  })
}

# Run CCA analyses
girona_cca_100m <- run_cca_analysis_100m(beaches_by_region$Girona, land_cover_data$Girona, "Girona")
barcelona_cca_100m <- run_cca_analysis_100m(beaches_by_region$Barcelona, land_cover_data$Barcelona, "Barcelona")
tarragona_cca_100m <- run_cca_analysis_100m(beaches_by_region$Tarragona, land_cover_data$Tarragona, "Tarragona")

# Combined analysis
all_species_data <- bind_rows(
  beaches_by_region$Girona %>% mutate(region = "Girona"),
  beaches_by_region$Barcelona %>% mutate(region = "Barcelona"),
  beaches_by_region$Tarragona %>% mutate(region = "Tarragona")
)

all_land_cover <- bind_rows(
  land_cover_data$Girona %>% mutate(region = "Girona"),
  land_cover_data$Barcelona %>% mutate(region = "Barcelona"),
  land_cover_data$Tarragona %>% mutate(region = "Tarragona")
)

combined_cca_100m <- run_cca_analysis_100m(all_species_data, all_land_cover, "Combined")

# Function to create CCA biplots
create_cca_biplot_100m <- function(cca_analysis, title_suffix = "") {
  if(is.null(cca_analysis)) return(NULL)
  
  cca_result <- cca_analysis$cca
  
  site_scores <- data.frame(
    CCA1 = scores(cca_result, display = "sites")[, 1],
    CCA2 = scores(cca_result, display = "sites")[, 2],
    Beach = cca_analysis$beach_ids
  )
  
  species_scores <- data.frame(
    CCA1 = scores(cca_result, display = "species")[, 1],
    CCA2 = scores(cca_result, display = "species")[, 2],
    Species = rownames(scores(cca_result, display = "species"))
  )
  
  env_scores <- data.frame(
    CCA1 = scores(cca_result, display = "bp")[, 1],
    CCA2 = scores(cca_result, display = "bp")[, 2],
    Variable = rownames(scores(cca_result, display = "bp"))
  )
  
  species_scores$distance <- sqrt(species_scores$CCA1^2 + species_scores$CCA2^2)
  top_species <- species_scores %>%
    arrange(desc(distance)) %>%
    slice_head(n = 8)
  
  p <- ggplot() +
    geom_point(data = site_scores, aes(x = CCA1, y = CCA2), 
               color = "blue", alpha = 0.6, size = 5) +
    geom_segment(data = env_scores, 
                 aes(x = 0, y = 0, xend = CCA1 * 2, yend = CCA2 * 2),
                 arrow = arrow(length = unit(0.8, "cm")), 
                 color = "red", size = 2, alpha = 0.8) +
    ggrepel::geom_text_repel(data = env_scores, 
              aes(x = CCA1 * 2.3, y = CCA2 * 2.3, label = Variable),
              color = "red", fontface = "bold", size = 8,
              max.overlaps = Inf) +
    geom_point(data = top_species, aes(x = CCA1, y = CCA2), 
               color = "darkgreen", alpha = 0.8, size = 4) +
    ggrepel::geom_text_repel(data = top_species, 
              aes(x = CCA1, y = CCA2, label = Species),
              color = "darkgreen", fontface = "italic", size = 6,
              max.overlaps = Inf) +
    theme_minimal() +
    labs(
      title = paste("CCA Biplot (100m scale):", cca_analysis$region, title_suffix),
      subtitle = paste("Explained variance:", round(cca_analysis$explained_variance, 1), "%"),
      x = paste("CCA1 (", round(100 * eigenvals(cca_result)[1] / sum(eigenvals(cca_result)), 1), "%)", sep = ""),
      y = paste("CCA2 (", round(100 * eigenvals(cca_result)[2] / sum(eigenvals(cca_result)), 1), "%)", sep = "")
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 24),
      plot.subtitle = element_text(size = 18),
      axis.title.x = element_text(size = 16),
      axis.title.y = element_text(size = 16),
      panel.grid.minor = element_blank(),
      aspect.ratio = 1
    )
  
  return(p)
}

# Test significance
test_env_significance_100m <- function(cca_analysis) {
  if(is.null(cca_analysis)) return(NULL)
  
  model_test <- anova(cca_analysis$cca, permutations = 999)
  var_test <- anova(cca_analysis$cca, by = "terms", permutations = 999)
  
  return(list(
    model = model_test,
    variables = var_test
  ))
}

girona_sig_100m <- test_env_significance_100m(girona_cca_100m)
barcelona_sig_100m <- test_env_significance_100m(barcelona_cca_100m)
tarragona_sig_100m <- test_env_significance_100m(tarragona_cca_100m)
combined_sig_100m <- test_env_significance_100m(combined_cca_100m)
```

This section presents Canonical Correspondence Analysis (CCA) results using land cover variables measured at **100m distance** from coastal dune sites. This broader spatial scale captures wider landscape context that may influence vegetation patterns beyond the immediate vicinity of the dunes.

**Scale Considerations**: The 100m scale is particularly relevant for understanding:
- Regional landscape influences on local vegetation patterns
- Connectivity effects between coastal and inland habitats  
- Broader disturbance gradients that operate at landscape scales
- Edge effects from land uses that may not directly border the beach

```{r}
#| label: fig-cca-combined-100m
#| fig-cap: "CCA biplot at 100m scale showing vegetation-landscape relationships across all coastal dune sites. Blue points represent beaches, red arrows show environmental gradients at 100m scale, and green points indicate key species."
#| fig-width: 20
#| fig-height: 20

if(!is.null(combined_cca_100m)) {
  p_combined_100m <- create_cca_biplot_100m(combined_cca_100m, "- All Regions")
  print(p_combined_100m)
  ggsave("../figures/cca_combined_analysis_100m.png", p_combined_100m, width = 12, height = 8)
} else {
  cat("Combined CCA analysis at 100m scale could not be performed.\n")
}
```

```{r}
#| label: fig-cca-regional-100m  
#| fig-cap: "Regional CCA biplots at 100m scale showing how broader landscape-vegetation relationships vary across the three coastal regions."
#| fig-width: 15
#| fig-height: 45

regional_plots_100m <- list()

if(!is.null(girona_cca_100m)) {
  regional_plots_100m$girona <- create_cca_biplot_100m(girona_cca_100m)
}
if(!is.null(barcelona_cca_100m)) {
  regional_plots_100m$barcelona <- create_cca_biplot_100m(barcelona_cca_100m)
}
if(!is.null(tarragona_cca_100m)) {
  regional_plots_100m$tarragona <- create_cca_biplot_100m(tarragona_cca_100m)
}

if(length(regional_plots_100m) > 0) {
  valid_plots_100m <- regional_plots_100m[!sapply(regional_plots_100m, is.null)]
  
  if(length(valid_plots_100m) > 0) {
    if(length(valid_plots_100m) == 3) {
      combined_regional_100m <- valid_plots_100m$girona / valid_plots_100m$barcelona / valid_plots_100m$tarragona
    } else if(length(valid_plots_100m) == 2) {
      combined_regional_100m <- valid_plots_100m[[1]] / valid_plots_100m[[2]]
    } else {
      combined_regional_100m <- valid_plots_100m[[1]]
    }
    
    print(combined_regional_100m)
    ggsave("../figures/cca_regional_analysis_100m.png", combined_regional_100m, width = 14, height = 12)
  }
}
```

```{r}
#| label: tbl-cca-summary-100m
#| tbl-cap: "Summary of CCA results at 100m scale showing variance explained by broader landscape variables."

cca_summary_100m <- data.frame(
  Region = character(),
  Explained_Variance_Percent = numeric(),
  Num_Environmental_Variables = integer(),
  Num_Species = integer(),
  Model_F_Value = numeric(),
  Model_P_Value = numeric(),
  stringsAsFactors = FALSE
)

extract_sig_info_100m <- function(sig_test) {
  if(is.null(sig_test) || is.null(sig_test$model)) {
    return(list(f_val = NA, p_val = NA))
  }
  return(list(
    f_val = sig_test$model$F[1],
    p_val = sig_test$model$`Pr(>F)`[1]
  ))
}

regions_100m <- list(
  list(name = "Girona", cca = girona_cca_100m, sig = girona_sig_100m),
  list(name = "Barcelona", cca = barcelona_cca_100m, sig = barcelona_sig_100m),
  list(name = "Tarragona", cca = tarragona_cca_100m, sig = tarragona_sig_100m),
  list(name = "Combined", cca = combined_cca_100m, sig = combined_sig_100m)
)

for(region in regions_100m) {
  if(!is.null(region$cca)) {
    sig_info <- extract_sig_info_100m(region$sig)
    
    cca_summary_100m <- rbind(cca_summary_100m, data.frame(
      Region = region$name,
      Explained_Variance_Percent = round(region$cca$explained_variance, 1),
      Num_Environmental_Variables = ncol(region$cca$env_matrix),
      Num_Species = ncol(region$cca$species_matrix),
      Model_F_Value = round(sig_info$f_val, 2),
      Model_P_Value = round(sig_info$p_val, 3),
      stringsAsFactors = FALSE
    ))
  }
}

if(nrow(cca_summary_100m) > 0) {
  knitr::kable(cca_summary_100m, 
               col.names = c("Region", "Explained Variance (%)", "Env. Variables", 
                           "Species", "F-value", "P-value"),
               align = c("l", "c", "c", "c", "c", "c"))
}
```

## **Multi-Scale Analysis: Comparing 50m vs 100m Results**

The 100m scale analysis provides complementary insights to the 50m analysis, revealing how vegetation patterns are influenced by environmental gradients operating at different spatial scales.

### **Scale-Dependent Environmental Drivers**

**Local Scale (50m) Effects:**
- Capture immediate environmental influences from adjacent land uses
- Reveal direct edge effects and microhabitat conditions
- Show relationships with proximate disturbance sources

**Landscape Scale (100m) Effects:**  
- Reveal broader landscape context influences
- Capture regional connectivity patterns and matrix effects
- Show cumulative impacts of multiple land use types

### **Conservation and Management Implications**

**Multi-Scale Conservation Strategy:**

1. **Local Management (50m focus):**
   - Direct habitat protection and restoration
   - Immediate threat mitigation
   - Buffer zone establishment around high-value areas

2. **Landscape Management (100m focus):**
   - Regional land use planning and zoning
   - Habitat connectivity corridor design
   - Coordination between multiple jurisdictions

3. **Integrated Approach:**
   - Species responding at both scales need management at both levels
   - Adaptive management considering scale-dependent responses
   - Monitoring programs designed to track changes at appropriate scales

### **Regional Patterns Across Scales**

Each coastal region shows different patterns of scale-dependence:

**Girona:** May show stronger landscape-scale effects due to more natural, connected habitats
**Barcelona:** Urban-natural gradients likely important at both scales  
**Tarragona:** Agricultural matrix effects may dominate at 100m scale

### **Future Research Directions**

1. **Temporal dynamics:** How do scale effects change over time?
2. **Functional traits:** Which species characteristics predict scale-dependent responses?
3. **Restoration effectiveness:** How do restoration outcomes vary by management scale?
4. **Climate change vulnerability:** Scale-dependent habitat requirements under changing conditions

This comprehensive multi-scale CCA analysis provides the scientific foundation for evidence-based coastal dune conservation that addresses environmental pressures operating across multiple spatial scales.
